Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package

### 2021-10-07 article_Bigsimr/article_Bigsimr.tex
rewriting as a single rmd to simplify pdf production

% Pandoc header
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{setspace}

@@

\hypertarget{background}{%
\section{Background}\label{background}}

The \texttt{Bigsimr} package presented here provides multiple high-performance algorithms that operate with HD multivariate data. All these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, \(B\), of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) \(F_i, i=1,\ldots,d\).

When designing this methodology, we developed the following properties to guide our effort. We divide the properties into two categories: (1) basic properties (BP) and `scalability' properties (SP). The BPs are adapted from an existing criteria due to \citet{Nik13a}. A suitable simulation strategy should possess the following properties:

\setstretch{1.5}

\begin{itemize}
\tightlist
\item
  BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
\item
  BP2: Flexible dependence, meaning that the number of dependence parameters can be equal to the number of bivariate pairs in the vector.
\item
  BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete marginal distributions.
\end{itemize}

\setstretch{2.0}

Moreover, the simulation method must \emph{scale} to high dimensions:

\setstretch{1.5}

\begin{itemize}
\tightlist
\item
  SP1: Procedure must scale to high dimensions with practical computation times.
\item
  SP2: Procedure must scale to high dimensions while maintaining accuracy.
\end{itemize}

\setstretch{2.0}

\hypertarget{motivating-example-rna-seq-data}{%
\subsection{Motivating example: RNA-seq data}\label{motivating-example-rna-seq-data}}

Simulating HD, non-normal, correlated data motivates this work, in pursuit of modeling RNA-sequencing (RNA-seq) data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients. The RNA-seq data-generating process involves counting how often a particular form of messenger RNA (mRNA) is expressed in a biological sample. RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in HD data. For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted \citep{Schissler2019}. Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes \citep{BE07, Schissler2018}.

We illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments). For ease of modeling and simplicity of exposition, we only consider high expressing genes. In turn, we begin by filtering to retain the top 1000 of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from \(N=878\) patients' tumor samples. This gives a great number of pairwise dependencies among the marginals (specifically, \(\ensuremath{4.995\times 10^{5}}\) correlation parameters). Table \ref{tab:ch010-realDataTab} displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples. To help visualize the bivariate relationships for these three selected genes across all patients, Figure \ref{fig:ch010-realDataFig} displays the marginal distributions and estimated Spearman's correlations.

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Patient ID & STAU1 & FKBP1A & NME2\\
\midrule
TCGA-A1-A0SB & 10440 & 11354 & 17655\\
TCGA-A1-A0SD & 21523 & 20221 & 14653\\
TCGA-A1-A0SE & 21733 & 22937 & 35251\\
TCGA-A1-A0SF & 11866 & 19650 & 16551\\
TCGA-A1-A0SG & 12486 & 12089 & 10434\\
\bottomrule
\end{tabular}
\end{table}

\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch010-realDataFig-1}

\caption[Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed]{Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed. The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology.}\label{fig:ch010-realDataFig}
\end{figure}
\end{CodeChunk}

\hypertarget{measures-of-dependency}{%
\subsection{Measures of dependency}\label{measures-of-dependency}}

In multivariate analysis, an analyst must select a metric to quantify dependency. The most widely-known is the Pearson correlation coefficient that describes the linear association between two random variables \(X\) and \(Y\), and, it is given by

\begin{equation}
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
\label{eq:pearson}
\end{equation}

As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal \((X,Y)\) random vector, the Pearson correlation completely describes the dependency between the components. For non-normal marginals with monotone correlation patterns, however, \(\rho_P\) suffers some drawbacks and may mislead or fail to capture important relationships \citep{MK01}. Alternatively, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted \(\rho_S\)) and Kendall's \(\tau\). Spearman's \(\rho_S\) has an appealing correspondence as the Pearson correlation coefficient on \emph{ranks} of the values, thereby capturing nonlinear yet monotone relationships. Kendall's \(\tau\), on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, \((X_i, Y_i)\) and \((X_j, Y_j)\) (concordance meaning that orderings have the same direction, e.g., if \(X_i < X_j\), then \(Y_i < Y_j\)). Note that concordance is determined by the ranks of the values, not the values themselves. Both \(\tau\) and \(\rho_S\) are \emph{invariant under monotone transformations} of the underlying random variates. As we will describe more fully in the \protect\hyperlink{algorithms}{Algorithms} section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).

\emph{Correspondence among Pearson, Spearman, and Kendall correlations}

There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions \(F_i\) are intrinsic in their calculation. For \emph{bivariate normal vectors}, however, the correspondence is well-known:

\begin{equation}
\label{eq:convertKendall}
\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's \(\rho\) \citep{K58},

\begin{equation}
\label{eq:convertSpearman}
\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

\emph{Marginal-dependent bivariate correlation bounds}

Given two marginal distributions, \(\rho_P\) is not free to vary over the entire range of possible correlations \([-1,1]\). The \emph{Frechet-Hoeffding bounds} are well-studied \citep{Nelsen2007, BF17}. These constraints cannot be overcome through algorithm design. In general, the bounds are given by

\begin{equation}
\label{eq:frechet}
\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, defined as

\begin{equation}
F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}
\label{eq:inverseCDF}
\end{equation}

when the variables are discrete.

\hypertarget{gaussian-copulas}{%
\subsection{Gaussian copulas}\label{gaussian-copulas}}

There is a connection of our simulation strategy to Gaussian \emph{copulas} \citep{Nelsen2007}. A copula is a distribution function on \([0,1]^d\) that describes a multivariate probability distribution with standard uniform marginals. This provides a powerful, natural way to characterize joint probability. Consequently, the study of copulas is an important and active area of statistical theory and practice. For any random vector \({\bf X}=(X_1, \ldots, X_d)^\top\) with CDF \(F\) and marginal CDFs \(F_i\) there is a copula function \(C(u_1, \ldots, u_d)\) satisfying

\begin{equation}
F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \quad x_i \in {\mathbb R}, i=1,\ldots,d.
\label{eq:copula}
\end{equation}

A Gaussian copula has marginal CDFs that are all standard normal, \(F_i = \Phi, \forall \, i\), . This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix \(R_P\). Since the marginals are standardized to have unit variance, this \(R_P\) is also Pearson correlation matrix. If \(F_R\) is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula \(C_R\) is defined through

\begin{equation}
\label{eq:gauss}
F_R(x_1, \ldots, x_d) = C_R(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

where \(\Phi(\cdot)\) is the standard normal CDF.

Here, the copula function \(C_R\) is the CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).

Sklar's Theorem \citep{Ubeda-Flores2017} guarantees that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with \(Y_i \sim F_i\), viz. \(Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).

\hypertarget{other-multivariate-simulation-packages}{%
\subsection{Other Multivariate Simulation Packages}\label{other-multivariate-simulation-packages}}

To place \texttt{Bigsimr}'s implementation and algorithms in context of existing approaches, we qualitatively compare \texttt{Bbigsimr} to other multivariate simulation packages. In \texttt{R}, the \texttt{MASS} package's \texttt{mvnorm()} has long been available to produce random vectors from multivariate normal (and \(t\)) distributions. This procedure cannot simulate from arbitrary margins and dependency measures and it fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy \texttt{random.multivariate\_normal} both do similar tasks. Recently, the high-performance \texttt{R} package \texttt{mvnfast} \citep{Fasiolo2016} has been released, which provides HD multivariate normal generation. However, lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD \texttt{R} packages exist, such as \texttt{GenOrd}. For copula computation, the full-featured \texttt{R} package \texttt{copula} provides many different copulas and a fairly general LD random generator \texttt{rCopula()}. High-dimesional random vector generation via \texttt{copula} takes an impractical amount of computing time \citep{Li2019gpu}. The \texttt{nortaRA} R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other \texttt{R} package that can produce HD random vectors with flexible margins and dependence structure. Table \ref{tab:compare-table} summarizes the above discussion.

\begin{table}[h]
\centering
\caption{\label{tab:compare-table}Comparison of selected features of \pkg{Bigsimr} to existing packages for multivariate simulation.}
\begin{tabular}{@{}llll@{}}
Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

\hypertarget{algorithms}{%
\section{Algorithms}\label{algorithms}}

This section describes our methods for simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\). Each \(Y_i\) has a specified marginal CDF \(F_i\) and its inverse \(F^{-1}_i\). To characterize dependency, every pair \((Y_i, Y_j)\) has a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\), and/or Kendall's \(\tau\). The method can be described as a \emph{high-performance Gaussian copula} (Equation \eqref{eq:gauss}) providing a HD NORTA-inspired algorithm.

\hypertarget{normal-to-anything-norta}{%
\subsection{NORmal To Anything (NORTA)}\label{normal-to-anything-norta}}

The well-known NORTA algorithm \citep{Cario1997} simulates a random vector \(\bf Y\) with variance-covariance matrix \(\Sigma_{\bf Y}\). Specifically, NORTA algorithm proceeds as follows:

\setstretch{1.5}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulate a random vector \(\bf Z\) with \(d\) independent and identically distributed (iid) standard normal components.
\item
  Determine the input matrix \(\Sigma_{\bf Z}\) that corresponds with the specified output \(\Sigma_{\bf Y}\).
\item
  Produce a Cholesky factor \(M\) of \(\Sigma_{\bf Z}\) such that \(M M^{\prime}=\Sigma_{\bf Z}\).
\item
  Set \(X\) by \(X \gets MZ\).
\item
  \(\text{Return} \; {\bf Y} \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d\).
\end{enumerate}

\setstretch{2.0}

With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability to HD.

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly. In general, there is no convenient correspondence between the components of the input \(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\). Matching the correlations involves evaluating or approximating \(\binom{d}{2}\) integrals of the form

\begin{equation}
    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
    \label{eq:pearsonIntegralRelation}
\end{equation}

where \(\phi(\cdot)\) is the joint probability density function of two correlated standard normal variables. For HD simulation, exact evaluation may become too costly to enable practical simulation studies. Our remedy to enable HD Pearson matching in \texttt{bigsimr} is Hermite polynomial approximation of these \(\binom{d}{2}\) double integrals, as suggested by \citet{XZ19}.

\emph{NORTA in higher dimensions}

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated. Even in LD spaces (e.g., \(d=3\)), there exist valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined, the resultant matrix becomes non-positive definite. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed \emph{NORTA defective}.

\citet{GH02} conducted a study to estimate the probability of encountering NORTA defective matrices while increasing the dimension \(d\). They found that for what is now considered low-to-moderate dimensions (\(d \approx 20\)), almost \emph{all} feasible matrices are NORTA defective. This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

There is hope, however, as \citet{GH02} also showed that replacing an non-positive definite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\). This provides evidence that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.

\hypertarget{rand-vec-gen}{%
\subsection{Random vector generator}\label{rand-vec-gen}}

We now describe our algorithm to generate random vectors, which extends NORTA to HD via nearest correlation matrix replacement and provides rank-based dependency matching:

\setstretch{1.5}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mapping step, either

  \begin{itemize}
  \tightlist
  \item
    Convert the target Spearman correlation matrix \(R_S\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Kendall \(\tau\) matrix \(R_K\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Pearson correlation matrix to \(R_P\) to the corresponding, approximate MVN Pearson correlation \(R_X\).
  \end{itemize}
\item
  Check correlation matrix admissibility and, if not, compute the nearest correlation matrix.

  \begin{itemize}
  \tightlist
  \item
    Check that \(R_X\) is a correlation matrix, a positive definite matrix with 1's along the diagonal.
  \item
    If the mapping produced an \(R_X\) that is a correlation matrix, we describe the matrix as \textbf{admissible} in this scheme and retain \(R_X\). Otherwise,
  \item
    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm and the result will be approximate.
  \end{itemize}
\item
  Gaussian copula

  \begin{itemize}
  \tightlist
  \item
    Generate \({\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)\).
  \item
    Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\) viz. \(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
  \item
    Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\).
  \end{itemize}
\end{enumerate}

\emph{Step 1: Mapping step.} We first employ the closed-form relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for bivariate normal random variables via Equations \eqref{eq:convertKendall} and \eqref{eq:convertSpearman}, respectively (implemented as \texttt{cor\_covert()}). Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.

For the computationally expensive process to match Pearson correlations, we approximate Equation \eqref{eq:pearsonIntegralRelation} for all pairs of margins. To this end, we implement the approximation scheme introduced by \citep{XZ19}. Briefly, the many double integrals of the form in \eqref{eq:pearsonIntegralRelation} are approximated by weighted sums of Hermite polynomials. Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the approximation is computationally expensive. To remedy this, we further approximate discrete distributions by a continuous distribution using Generalized S-Distributions \citep{muino2006gs}.

\emph{Step 2: Admissibility check and nearest correlation matrix computation.} Once \(R_X\) has been determined, we check admissibility of the adjusted correlation via the steps described above. If \(R_X\) is not a valid correlation matrix, then we compute the nearest correlation matrix. Finding the nearest correlation matrix is a common statistical computing problem. The defacto function in \texttt{R} is \texttt{Matrix::nearPD}, an alternating projection algorithm due to \citet{higham2002computing}. As implemented, the function fails to scale to HD. Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions \citep{qi2006quadratically}. The nearest correlation matrix problem can be written down as the following convex optimization problem: \(\mathrm{min} \quad \frac{1}{2} \Vert R_X - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}\).

For nonparametric correlation measures, our algorithm allows the generation of HD multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall \(\tau\) matrices. The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN. In particular, if we let \(X\) be MVN with \(d\) components and denote \(\Omega_P = \{ R_P : R_P \textrm{ is a Pearson correlation matrix for } X \}, \quad \Omega_K = \{ R_K : R_K \textrm{ is a Kendall correlation matrix for } X \}, \quad \Omega_S = \{ R_S : R_S \textrm{ is a Spearman correlation matrix for } X \}\). There are 1-1 mappings between these sets. We conjecture that the sets of admissible \(R_S\) and \(R_K\) are not highly restrictive. In particular, \(R_P\) is approximately \(R_S\) for a MVN, suggesting that the admissible set \(\Omega_S\) should be flexible as \(R_P\) can be any PD matrix with 1's along the diagonal. We provide methods to check whether a target \(R_S\) is an element of the \emph{admissible set} \(\Omega_S\).

There is an increasing probability of encountering an inadmissible correlation matrix as dimension increases. In our experience, the mapping step for large \(d\) almost always produces a \(R_X\) that is not a correlation matrix. In Section \ref{package}, we provide a basic method of how to quantify and control the approximation error. Further, the RNA-seq example in Section \ref{examples} provides an illustration of this in practice.

\emph{Step 3: Gaussian copula.} The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins. Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation. Step 3 requires a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals. The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF \(\Phi(\cdot)\). Finally, the quantile functions \(F_i^{-1}\) are applied across the margins to return the desired random vector \({\bf Y}\).

\section[Bigsimr-pkg]{The \pkg{Bigsimr} package}\label{package}

This section describes a bivariate random vector simulation workflow via the \texttt{bigsimr} \texttt{R} package. This \texttt{R} package provides an interface to the native code written in Julia (registered as the \texttt{Bigsimr} Julia package). In addition to the native Julia \texttt{Bigsimr} package and \texttt{R} interface \texttt{bigsimr}, we also provide a Python interface \texttt{bigsimr} that interfaces with the Julia \texttt{Bigsimr} package. The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \ref{algorithms}).

The subsections below describe the basic use of the \texttt{bigsimr} R package by stepping through an example workflow using the data set \texttt{airquality} that contains daily air quality measurements in New York, May to September 1973 \citep{Chambers1983}. This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.

\hypertarget{bivariate-example}{%
\subsection{Bivariate example}\label{bivariate-example}}

We illustrate the use of \texttt{bigsimr} using the New York air quality data set (\texttt{airquality}) included in the R \texttt{datasets} package. First, we load the \texttt{bigsimr} library and a few other convenient data science packages, including the syntactically-elegant \texttt{tidyverse} suite of \texttt{R} packages. The code chunk below prepares the computing environment:

\begin{CodeChunk}
\begin{CodeInput}
R> library("tidyverse")
R> library("bigsimr")
R> # Activate multithreading in Julia
R> Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
R> # Load the Bigsimr and Distributions Julia packages
R> bs <- bigsimr_setup()
R> dist <- distributions_setup()
\end{CodeInput}
\end{CodeChunk}

Here, we describe a minimal working example --- a bivariate simulation of two airquality variables: \texttt{Temperature}, in degrees Fahrenheit, and \texttt{Ozone} level, in parts per billion.

\begin{CodeChunk}
\begin{CodeInput}
R> df <- airquality %>% select(Temp, Ozone) %>% drop_na()
R> glimpse(df)
\end{CodeInput}
\begin{CodeOutput}
Rows: 116
Columns: 2
$ Temp  <int> 67, 72, 74, 62, 66, 65, 59, 61, 74, 69, 66, 68, 58, 64, 66, 57, ~
$ Ozone <int> 41, 36, 12, 18, 28, 23, 19, 8, 7, 16, 11, 14, 18, 14, 34, 6, 30,~
\end{CodeOutput}
\end{CodeChunk}

Figure \ref{fig:ch030-aq-joint-dist} visualizes the bivariate relationship between \texttt{Ozone} and \texttt{Temperature}. We aim to simulate random two-component vectors mimicking this structure. The margins are not normally distributed; \texttt{Ozone} level exhibits a strong positive skew.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch030-aq-joint-dist-1}

\caption[Bivariate scatterplot of Ozone versus Temp with estimated marginal densities]{Bivariate scatterplot of Ozone versus Temp with estimated marginal densities. The Ozone data are modeled marginally as log-normal and the Temperature data as normal.}\label{fig:ch030-aq-joint-dist}
\end{figure}
\end{CodeChunk}

Next, we specify the marginal distributions and correlation coefficient (both type and magnitude). Here the analyst is free to be creative. For this example, we avoid goodness-of-fit considerations to determine the marginal distributions. It is sensible without domain knowledge to estimate these quantities from the data, and \texttt{bigsimr} contains fast functions designed for this task.

\hypertarget{specifying-marginal-distributions}{%
\subsection{Specifying marginal distributions}\label{specifying-marginal-distributions}}

Based on the estimated densities in Figure \ref{fig:ch030-aq-joint-dist}, we assume \texttt{Temp} is normally distributed and \texttt{Ozone} is log-normally distributed, as the latter values are positive and skewed. We use the classical unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

\begin{CodeChunk}
\begin{CodeInput}
R> df %>% select(Temp) %>% 
+   summarise_all(.funs = c(mean = mean, sd = sd))
\end{CodeInput}
\begin{CodeOutput}
   mean    sd
1 77.87 9.485
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> mle_mean <- function(x) mean(log(x))
R> mle_sd <- function(x) mean( sqrt( (log(x) - mean(log(x)))^2 ) )
R> df %>% 
+   select(Ozone) %>% 
+   summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
\end{CodeInput}
\begin{CodeOutput}
  meanlog  sdlog
1   3.419 0.6967
\end{CodeOutput}
\end{CodeChunk}

Now, we configure the input marginals for later input into \texttt{rvec}. The marginal distributions are specified using \texttt{Julia}'s \texttt{Distributions} package and stored in a vector.

\begin{CodeChunk}
\begin{CodeInput}
R> margins <- c(dist$Normal(mean(df$Temp), sd(df$Temp)),
+              dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone)))
\end{CodeInput}
\end{CodeChunk}

\hypertarget{specifying-correlation}{%
\subsection{Specifying correlation}\label{specifying-correlation}}

The user must decide how to describe correlation based on the particulars of the problem. For non-normal data in our scheme, we advocate the use of Spearman's \(\rho\) correlation matrix \(R_S\) or Kendall's \(\tau\) correlation matrix \(R_K\). We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for the distribution at hand (see \protect\hyperlink{simulations}{Monte Carlo evaluations} below for evaluation strategies and guidance). Note that these estimation methods are classical approaches, not specifically designed for high-dimensional correlation estimation.

\begin{CodeChunk}
\begin{CodeInput}
R> (R_S <- bs$cor(as.matrix(df), bs$Spearman))
\end{CodeInput}
\begin{CodeOutput}
      [,1]  [,2]
[1,] 1.000 0.774
[2,] 0.774 1.000
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{checking-target-correlation-matrix-admissibility}{%
\subsection{Checking target correlation matrix admissibility}\label{checking-target-correlation-matrix-admissibility}}

Once a target correlation matrix is specified, first convert the dependency type to Pearson to correctly specify the MVN inputs. Then, check that the converted matrix \(R_X\) is a valid correlation matrix (PD and 1s on the diagonal). For this bivariate example, \(R_X\) is a valid correlation matrix. Note that typically in HD the resultant MVN Pearson correlation matrix is indefinite and requires approximation (see the HD examples in subsequent sections).

\begin{CodeChunk}
\begin{CodeInput}
R> # Step 1. Mapping
R> (R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
\end{CodeInput}
\begin{CodeOutput}
       [,1]   [,2]
[1,] 1.0000 0.7886
[2,] 0.7886 1.0000
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> # Step 2. Check admissibility
R> bs$iscorrelation(R_X)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

Despite being a valid correlation matrix under this definition, marginal distributions induce Frechet bounds on the possible Pearson correlation values. To check that the converted correlation values fall within these bounds, use \texttt{bigsimr::cor\_bounds} to estimate the pairwise lower and upper correlation bounds. \texttt{cor\_bounds} uses the Generate, Sort, and Correlate algorithm of \citet{DH2011}. For the assumed marginals, the Pearson correlation coefficient is not free to vary in {[}-1,1{]}, as seen below. Our MC estimate of the bounds slightly underestimates the theoretical bounds of \((-0.881, 0.881)\). (See an analytic derivation presented in the Appendix). Since our single Pearson correlation coefficient is within the theoretical bounds, the correlation matrix is valid for our simulation strategy.

\begin{CodeChunk}
\begin{CodeInput}
R> bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
\end{CodeInput}
\begin{CodeOutput}
Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
(lower = -0.8823026550919685, upper = 0.8815987574226597)
\end{CodeOutput}
\end{CodeChunk}

In this 2D example, it is possible to verify the existence of a bivariate distribution with the target Pearson correlation / margins (See the \protect\hyperlink{appendix}{Appendix}).
For general \(d-\)variate distributions, is it not guaranteed that such a multivariate distribution exists with exact the specified correlations and margins \citep{BF17}.
Theoretical guidance is lacking in this regard, but, in practice, we find that using bivariate bounds form rough approximations to the bounds for the feasible region for the \(d-\)variate construction when aggregated into a Pearson matrix.
See the \protect\hyperlink{examples}{RNA-seq data application} section for an example discussing this, NORTA feasibility, and use of the nearest PD matrix algorithm.

\hypertarget{simulating-random-vectors}{%
\subsection{Simulating random vectors}\label{simulating-random-vectors}}

Finally, we execute \texttt{rvec} to simulate the desired \(10,000\) random vectors from the assumed bivariate distribution of \texttt{Ozone} and \texttt{Temp}. Note that the target input is specified using the pre-computed MVN correlation matrix. Figure \ref{fig:ch030-plot-sim} plots the 10,000 simulated points.

\begin{CodeChunk}
\begin{CodeInput}
R> x <- bs$rvec(10000, R_X, margins)
R> df_sim <- as.data.frame(x)
R> colnames(df_sim) <- colnames(df)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch030-plot-sim-1}

\caption[Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels]{Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association.}\label{fig:ch030-plot-sim}
\end{figure}
\end{CodeChunk}

\hypertarget{simulations}{%
\section{Monte Carlo evaluations}\label{simulations}}

In this section, we conduct MC studies to investigate method performance. Marginal parameter matching is straightforward in our scheme as it is a sequence of univariate inverse probability transforms. This process is both fast and exact. On the other hand, accurate and computationally efficient dependency matching presents a challenge, especially in the HD setting. To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension \(d\).

\hypertarget{bivariate-experiments}{%
\subsection{Bivariate experiments}\label{bivariate-experiments}}

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example. Thus, we proceed by increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data. We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations. For each pair of identical margins, we vary the target correlation across \(\Omega\), the set of possible admissible values for each correlation type, to evaluate the simulation's ability to generate all possible correlations. The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).

Table \ref{tab:sims} lists our identical-marginal, bivariate simulation configurations. We increase the simulation replicates \(B\) to visually ensure that our results converge to the target correlations and gauge efficiency. We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types. Then, we select a non-symmetric continuous distribution: a standard (rate =1), two-component multivariate gamma (MVG). Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see \href{examples}{RNA-seq data application} for estimation details). Specifically, we assume a MVNB with \(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).

>>>>

% Pandoc header

\usepackage{amsmath}

@@
